### 2019-03-01 submission/manuscript.tex

PCR amplification of 16S rRNA genes is a critical, yet under appreciated
step in the generation of sequence data to describe the taxonomic
composition of microbial communities. Numerous factors in the design of
PCR can impact the sequencing error rate, the abundance of chimeric
sequences, and the degree to which the fragments in the product
represent their abundance in the original sample. The choice of PCR
cycling conditions are known to impact the outcome of PCR and downstream
inferences. We compared the performance of high fidelity polymerases and
varying number of rounds of amplification when amplifying a mock
community with known sequences and community structure and human stool
samples. Although it was impossible to derive specific recommendations,
we did observe general trends. Namely, using a polymerase with the
highest possible fidelity and minimizing the number of rounds of PCR
reduced the sequencing error rate, fraction of chimeric sequences, and
bias. Evidence of bias at the sequence level was subtle and could not be
ascribed to the fragments' fraction of bases that were guanines or
cytosines. When analyzing mock community data, the amount that the
community deviated from the expected composition increased with rounds
of PCR. The effect was inconsistent for human stool samples. Overall the
results underscore the difficulty of comparing sequence data that are
generated by different PCR protocols. However, the results indicate that
the variation in human stool samples is generally larger than that
introduced by the choice of polymerase or number of rounds of PCR.

@@

16S rRNA gene sequencing is a powerful and widely used tool for
surveying the structure of microbial communities (1--3). This approach
has exploded in popularity with the advent of high throughput sequencing
where it is possible to characterize numerous samples with thousands of
sequences per sample. Numerous factors can impact how a natural
community is represented by the sequencing data including the method of
acquiring samples (4--8), storage conditions (4--6, 9--12), extraction
methods (13), amplification conditions (8, 14, 15), sequencing method
(15--17), and analytical pipeline (15, 18--20). The increased sampling
depth that is now available relative to previous Sanger sequencing-based
methods is expected to compound the impacts of an investigator's choices
and the interpretation of their results.

One step in the generation of 16S rRNA gene sequence data that has been
long known to have a significant impact on the description of microbial
communities is the choice of conditions for PCR amplification (8, 14,
15). Factors such as the choice of primers have an obvious impact on
which populations will be amplified (18, 21). However, a variety of PCR
artifacts can also impact the perception of a community including the
formation of chimeras (14, 22--24), misincorporation of nucleotides (23,
25, 26), preferential amplification of populations leading to bias (24,
27--33), and accumulation of random amplification events leading to PCR
drift (24, 27, 32, 34). Many bioinformatic tools have been developed to
identify chimeras, there are significant sensitivity and specificity
tradeoffs (14, 35). Laboratory-based solutions to minimize chimera
formation have also been proposed such as minimizing the amount of
template DNA in the PCR, minimizing the number of rounds of PCR,
minimizing the amount of shearing in the template DNA, and using DNA
polymerases that have a proof-reading ability (14, 23). Others have
attempted to account for PCR bias using modeling approaches (29, 36);
however, these have been developed for idealized situations. To minimize
PCR drift, some investigators will pool technical replicate PCRs hoping
to average out the drift (34). Other factors that have been shown to
impact the formation of PCR artifacts are outside the control of an
investigator including the fraction of DNA bases that are guanines or
cytosines, the variation in the length of the targeted region across the
community, the sequence of the DNA that flanks the template, and the
genetic diversity of the community (28, 30--33). Early investigations of
the factors that lead to the formation of PCR artifacts focused on
analyzing binary mixtures of genomic DNA and 16S rRNA gene fragments to
explore PCR biases and chimera formation. Although these studies were
instrumental in forcing researchers to be cautious about the
interpretation of their results, we have a poor understanding of how
these factors affect the formation of PCR artifacts in more complex
communities.

@@

This study investigated the impact of choice of high fidelity DNA
polymerase and the number of rounds of amplification on the formation of
PCR artifacts using a mock community and human stool samples. These PCR
artifacts included (i) the effect of the polymerase on the error rate of
the bases represented in the final sequences, (ii) the fraction of
sequences that appeared to be chimeras and the ability to detect those
chimeras, (iii) the bias of preferentially amplifying one fragment over
another in a mixed pool of templates, and (iv) inter-sample variation in
community structure of samples amplified with the same polymerase across
the amplification process. To characterize these factors we sequenced a
mock community of 8 organisms with known sequences and community
structure and human fecal samples with unknown sequences and community
structures. We sequenced the V4 region of the 16S rRNA genes from a mock
community by generating paired 250 nt reads on the Illumina MiSeq
platform. This region and sequencing approach was used because it has
been shown to result in a relatively low sequencing error rate and is a
widely used protocol ({\textbf{???}}). To better understand the impact
of DNA polymerase choice on PCR artifacts, we selected five high
fidelity DNA polymerases and amplified the communities using 20, 25, 30,
and 35 rounds of amplification. Collectively, our results suggest that
the number of rounds and to a lesser degree the choice of DNA polymerase
used in PCR impact the sequence data, the effects are consistent and are
smaller than the biological differences between individuals.

@@

\textbf{\emph{The fraction of sequences identified as being chimeric
varies by the number of cycles and the DNA polymerase used in PCR.}}
Chimeric PCR products can significantly confound downstream analyses.
Although numerous bioinformatic tools exist to identify and remove
chimeric sequences with high specificity, their sensitivity is
relatively low and can be reduced by the presence of sequencing errors
(14, 35). Because the true sequences of the organisms in the mock
community were known, we could generate all possible chimeras between
pairs of V4 16S rRNA gene fragments and use these possible chimeric
sequences to screen the sequences generated under the different PCR
conditions for chimeras. The number of chimeras increased with rounds of
amplification (Figure 2A). Interestingly, the fraction of chimeric
sequences from the mock community varied by the type of polymerase used.
After 30 rounds of PCR, the Platinum polymerase had the highest chimera
rate (i.e.~18.2\%) followed by the Q5 (i.e.~8.1\%), Phusion
(i.e.~7.5\%), Kappa (i.e.~2.3\%), and Accuprime (i.e.~0.9\%)
polymerases. Because our chimera screening procedure can only be applied
to mock communities, we used the UCHIME algorithm to model the chimera
screening approach that is used in most sequence curation pipelines. By
comparing the output of UCHIME to our reference-based approach, we were
able to calculate the UCHIME's sensitivity and specificity (Figure 2A).
The specificity for all polymerases was above 95.7\% and showed a weak
association with the number of cycles used. There was considerable
inter-polymerase and inter-round of amplification variation in the
sensitivity of UCHIME to detect the chimeras from the mock community.
This suggested that the residual error rate after pre-clustering the
sequence data did not compromise the sensitivity. The sensitivity of
UCHIME varied between 50 and 87.0\%. The generalizability of these
results is limited because we used a single mock community with limited
genetic diversity. Although we did not know the true chimera rate for
our four human stool samples, we were able to calculate the fraction of
sequences that UCHIME identified as being chimeric (Figure 2B). These
results followed those from the mock communities: additional rounds of
amplification significantly increased the rate of chimeras and there was
variation between the polymerases that we used. Although it was not
possible to identify the features of a polymerase that resulted in
higher rates of chimeras, it is clear that using the smallest number of
PCR cycles possible will minimize the impacts of chimeras.

\textbf{\emph{At the sequence level, PCR amplification bias is subtle.}}
Since researchers began using PCR to amplify 16S rRNA gene fragments
there has been concern that amplifying fragments from a mixed template
pool could lead to a biased representation in the pool of products and
would confound downstream analyses (24, 27--33). The mock community was
generated by mixing equal amounts of genomic DNA from each of the 8
bacteria resulting in uneven representation of the \emph{rrn} operons
across the bacteria because each bacterium had a different genome size
and varied in the number of operons in its genome. The vendor of the
mock community subjects each lot of genomic DNA to shotgun sequencing to
more accurately quantify the actual abundance of each organism in the
community. We compared the expected relative abundance of the 16S rRNA
genes from each bacterium in the mock community to the data we generated
across rounds of amplification and polymerase (Figure 3). Interestingly,
for some bacteria, their representation became less biased with
additional rounds of PCR (e.g.~\emph{L. fermentum}), while others became
more biased (e.g.~\emph{E. faecalis}), and others had little change
(e.g.~\emph{B. subtilis}). Contrary to prior reports (28), The
percentage of bases in the V4 region that were guanines or cytosines was
not predictive of the amount of bias. Across the strains there was no
variation in the length of their V4 regions and they each had the same
sequence in the region that the primers annealed. One of the bacteria
represented in the mock community, \emph{S. enterica}, had 6 identical
copies of the V4 region and 1 copy that differed from those by one
nucleotide. The dominant copy had a thymidine and the rare copy had a
guanine. We used the sequence data to calculate the ratio of the
dominant to rare variants from \emph{S. enterica} (Figure S1). The
Accuprime, Phusion, Platinum, and Q5 polymerases converged to a ratio of
5.4; however, the ratio for the Kappa polymerase varied between 6.1 and
6.8 for 25 to 35 rounds of PCR. Given the subtle nature of the variation
in the relative abundances of each 16S rRNA gene fragment, it was not
possible to create generalizable rules that would explain the bias.

\textbf{\emph{At the community level, the effects of PCR amplification
bias grow with additional rounds of PCR.}} Because the variation in bias
between polymerases and across rounds of PCR could be artificially
inflated due to sequencing errors and chimeras, we analyzed the alpha
and beta diversity of the mock community data at different phases of the
sequence curation pipeline (Figure 4). First, we removed the chimeras
from the mock community data as described above and mapped the
individual reads to the OTUs that the 16S rRNA gene fragments would
cluster into if there were no sequencing errors. This gave us a
community distribution that reflected the distribution following PCR
without any artifacts. Although the richness did not change, the Shannon
diversity increased with the number of rounds of PCR for all polymerases
except for the Kappa polymerase, for which the diversity decreased
(Figure 4A). These data suggest that PCR had the effect of making the
community distribution more even than it was originally, except for the
data generated using the Kappa polymerase where the evenness decreased.
Next, we used the observed sequence errors, but removed chimeras using
our reference-based approach, and clustered the reads to OTUs. The
richness and diversity metrics trended higher with higher error rates
and number of rounds of PCR (Figure 4A). Finally, we used the observed
sequence data and used the UCHIME algorithm to identify chimeras. Again,
the richness and diversity metrics trended higher with higher error
rates and number of rounds of PCR (Figure 4A). These comparisons
demonstrated that although the bias at the sequence level was subtle,
PCR introduces bias that is exacerbated by errors and chimeras when
sequences are clustered into OTUs. When we measured the Bray-Curtis
distance between the communities observed after 25 rounds of
amplification and those at 30 and 35 rounds of amplification the
distances between 25 and 35 rounds were higher than between 25 and 30
rounds for each of the polymerases by an average of 0.022 units (Figure
4B). The Platinum polymerase varied the most across rounds of
amplification (25 vs 30 rounds: 0.12; 25 vs 35 rounds: 0.15). Although
the distances between samples were small, the ordination of these
distances showed a clear change in community structure with increasing
rounds of PCR (Figure 4C). This observation was supported by our
statistical analysis, which revealed that the number of rounds of PCR
(R\textsuperscript{2}=0.21, P\textless{}0.001) was a larger factor than
the choice of polymerase (R\textsuperscript{2}=0.20, P\textless{}0.001).
These results demonstrate that subtle differences in relative abundances
can have an impact on overall community structure. This variation
underscores the importance of only comparing sequence data that have
been generated using the same PCR conditions.

\textbf{\emph{The choice of polymerase or the number of rounds of
amplification have little impact on the relative interpretation of
community-wide metrics of diversity.}} The biases that we observed at
the population and community levels using mock community data appeared
to be small relative to the expected differences between biological
samples. To study this further, we calculated alpha and beta-diversity
metrics using the human stool samples for each of the polymerases and
rounds of amplification. We calculated the number of observed OTUs and
Shannon diversity for each condition and subject (Figure 5A). Although
there were clear differences between conditions, the relative ordering
of the stool samples did not meaningfully vary across conditions. When
we characterized the variation between rounds of amplification using
human stool samples, the distance between the 25 and 30 rounds and 25
and 35 rounds varied considerably between samples and polymerases
(Figure 5B). In general the inter-round variation was lowest for the
data generated using the Kappa and Accuprime polymerases. The data
generated using the Platinum polymerase was consistent across rounds,
but overall, it was more biased than the other polymerases. Considering
the average distance across the four samples varied between 0.39 and
0.56, regardless of the polymerases and number of rounds of
amplification, any bias due to amplification is unlikely to obscure
community-wide differences between samples. In support of this was our
principle coordinates analysis of the Bray-Curtis distances, which
revealed distinct clusters by donor (Figure 5C). Within each cluster
there were no obvious patterns related to the polymerase or number of
rounds of PCR. However, our statistical analysis revealed statistically
significant differences in the community structures with the subject
explaining the most variation (R\textsuperscript{2}=0.79,
P\textless{}0.001), followed by the number of rounds of PCR
(R\textsuperscript{2}=0.044, P\textless{}0.001) and the choice of
polymerase (R\textsuperscript{2}=0.033, P\textless{}0.001). Together,
these results indicate that for a coarse analysis of communities, the
choice of number of rounds of amplification or polymerase are not
important, but that they must be consistent across samples. It is
difficult to develop a specific recommendation based on the level of
bias across rounds of PCR or polymerases; however, the general
suggestion is to use as few rounds of amplification as possible.

@@

Our results suggest that the number of rounds of PCR and to a lesser
degree the choice of DNA polymerase impact the analysis of 16S rRNA gene
sequence data from bacterial communities. Although it was not possible
to make direct connections between PCR conditions and specific sources
of bias, we were able to identify general recommendations that reduce
the amount of error, chimera formation, and bias. Researchers should
strive to minimize the number of rounds of PCR and should use a high
fidelity polymerase. Although specific PCR conditions impact the precise
interpretation of the data, the effects were consistent and were smaller
than the biological differences between individuals. Based on these
observations, amplicons must be generated by consistent protocols to
yield meaningful comparisons. When comparing across studies, values like
richness, diversity, and relative abundances must be made in relative
and not absolute terms.

The observed sequencing error rates and alpha diversity metrics followed
the manufacturers' measurements of their polymerases' fidelity (Figure
1). Accuprime and Platinum have fidelity that are approximately 10-times
higher than that of Taq whereas the fidelity of Phusion, Q5, and Kappa
are more than 100 times higher. Among these polymerases, the Kappa
polymerase resulted in the the lowest error rate, lowest chimera rate,
and least bias across rounds of PCR. Considering polymerase development
is an active area of commercial development and it likely that new
polymerases will come to market, it is important for researchers to
understand how changing the polymerase impacts downstream analyses for
their type of samples.

Over the past 20 years, a large literature has attempted to document
various PCR biases and underscores the fact that data based on
amplification of DNA from a mixed community are not a true
representation of the actual community. In addition to obvious biases
imposed by primer selection, other factors inherent in PCR can influence
the representation of communities. Factors that can lead to preferential
amplification of one fragment over another have included guanine and
cytosine composition, length, flanking DNA composition, amount of DNA
shearing, and number of rounds of PCR (24, 27--33). In addition,
environmental and reagent contaminants can also have a significant
impact on the analysis of low biomass samples (39). Less well understood
is the effect of phylogenetic diversity on bias and chimera formation.
Communities with low phylogenetic diversity may be more prone to chimera
formation since chimeras are more likely to form among closely related
sequences (14, 35). The interaction of these various influences on PCR
artifacts are complex and difficult to tease apart. Minimizing the level
of DNA shearing and using the fewest number of rounds of PCR with a
polymerase that has the highest possible fidelity are strategies that
can be employed to minimize the formation of chimeras. Although care
should always be taken when choosing a polymerase for 16S rRNA gene
sequencing, our observations show that the differences between a variety
of polymerases are smaller than the actual biological variation in fecal
communities between individuals. Therefore, if the biological signal of
interest is similar to differences in fecal bacterial communities found
between individuals, then the type of high fidelity DNA polymerase used
will only minimally change the results.

Even with these strategies it is impossible to remove all PCR artifacts.
Beyond the imperfections of the best polymerases, sometimes difficult to
lyse organisms require stringent lysis steps and low biomass samples
require additional rounds of PCR. A host of bioinformatics tools are
available for removing residual sequencing errors (18, 40--42). These
tools struggle to correctly differentiate true biological diversity
(i.e.~Amplicon Sequencing Variants or ASVs) and PCR or sequencing
errors. Although these methods remove residual errors, they also risk
splitting intragenomic variants into separate ASVs or merging 16S rRNA
gene sequences from different taxa into the same ASV. Other tools are
available for removing chimeras (14, 35) where there is a trade off
between the sensitivity of detecting chimeras and the specificity of
correctly calling a sequence a chimera. In recent years, parameters for
these algorithms have been changed to increase their sensitivity with
little evaluation of the effects on the specificity of the algorithms
(40, 42). Others recommend removing any read that has an abundance below
a specified threshold as a tool to remove PCR and sequencign artifacts
(e.g.~removing all sequences that only appear once) (20, 40--42). This
method must be approached with caution as such approaches are likely to
introduce a different bias of the community representation and ignore
the fact that artifacts may be quite abundant. Ultimately, researchers
must test their hypotheses with multiple methods to validate the claims
they reach with any one method (43). All methods have biases and
limitations and we must use complementary methods to obtain robust
results.

@@

\textbf{\emph{PCR protocol.}} Five high fidelity DNA polymerases were
tested including AccuPrime\textsuperscript{TM} (ThermoFisher, MA, USA),
KAPA HIFI (Roche, IN, USA), Phusion (New England Biolabs, MA, USA),
Platinum (ThermoFisher, MA, USA), and Q5 (New England Biolabs, MA, USA).
Manufacturer recommendations were followed except for the annealing and
extension times, which were selected based on previously published
protocols (18, 37). Primers targeting the V4 region of the 16S rRNA gene
were used with modifications to generate MiSeq amplicon libraries (18)
(\url{https://github.com/SchlossLab/MiSeq_WetLab_SOP/}). The number of
rounds of PCR used for each sample and polymerase 15 and increased by 5
rounds up to 35 cycles. Insufficient PCR product was generated using 15
rounds and has not been included in our analysis.

>>>>

PCR amplification of 16S rRNA genes is a critical, yet under appreciated
step in the generation of sequence data to describe the taxonomic
composition of microbial communities. Numerous factors in the design of
PCR can impact the sequencing error rate, the abundance of chimeric
sequences, and the degree to which the fragments in the product
represent their abundance in the original sample (i.e.~bias). We
compared the performance of high fidelity polymerases and varying number
of rounds of amplification when amplifying a mock community and human
stool samples. Although it was impossible to derive specific
recommendations, we did observe general trends. Namely, using a
polymerase with the highest possible fidelity and minimizing the number
of rounds of PCR reduced the sequencing error rate, fraction of chimeric
sequences, and bias. Evidence of bias at the sequence level was subtle
and could not be ascribed to the fragments' fraction of bases that were
guanines or cytosines. When analyzing mock community data, the amount
that the community deviated from the expected composition increased with
rounds of PCR. This bias was inconsistent for human stool samples.
Overall the results underscore the difficulty of comparing sequence data
that are generated by different PCR protocols. However, the results
indicate that the variation in human stool samples is generally larger
than that introduced by the choice of polymerase or number of rounds of
PCR.

@@

16S rRNA gene sequencing is a powerful and widely used tool for
surveying the structure of microbial communities (1--3). This approach
has exploded in popularity with the advent of high throughput sequencing
where it is possible to characterize numerous samples with thousands of
sequences per sample. Many factors can impact how a natural community is
represented by the sequencing data including the method of acquiring
samples (4--8), storage conditions (4--6, 9--12), extraction methods
(13), amplification conditions (8, 14, 15), sequencing method (15--17),
and analytical pipeline (15, 18--20). The increased sampling depth that
is now available relative to previous Sanger sequencing-based methods is
expected to compound the impacts of an investigator's choices and the
interpretation of their results.

One step in the generation of 16S rRNA gene sequence data that has been
long known to have a significant impact on the description of microbial
communities is the choice of conditions for PCR amplification (8, 14,
15). Factors such as the choice of primers have an obvious impact on
which populations will be amplified (18, 21). However, a variety of PCR
artifacts can also impact the perception of a community including the
formation of chimeras (14, 22--24), misincorporation of nucleotides (23,
25, 26), preferential amplification of some populations over others
leading to bias (24, 27--33), and accumulation of random amplification
events leading to PCR drift (24, 27, 32, 34). Many bioinformatic tools
have been developed to identify chimeras; however, there are significant
sensitivity and specificity tradeoffs (14, 35). Laboratory-based
solutions to minimize chimera formation have also been proposed such as
minimizing the amount of template DNA in the PCR, minimizing the number
of rounds of PCR, minimizing the amount of shearing in the template DNA,
and using DNA polymerases that have a proof-reading ability (14, 23).
Others have attempted to account for PCR bias using modeling approaches
(29, 36). In cases where there has been success it has been with
relatively small communities with consistent composition (29). To
minimize PCR drift, some investigators pool technical replicate PCRs
hoping to average out the drift (34). Other factors that have been shown
to impact the formation of PCR artifacts are outside the control of an
investigator including the fraction of DNA bases that are guanines or
cytosines, the variation in the length of the targeted region across the
community, the sequence of the DNA that flanks the template, and the
genetic diversity of the community (28, 30--33). Early investigations of
the factors that lead to the formation of PCR artifacts focused on
analyzing binary mixtures of genomic DNA and 16S rRNA gene fragments to
explore PCR biases and chimera formation. Although these studies were
instrumental in forcing researchers to be cautious about the
interpretation of their results, we have a poor understanding of how
these factors affect the formation of PCR artifacts in more complex
communities.

@@

This study investigated the impact of choice of high fidelity DNA
polymerase and the number of rounds of amplification on the formation of
PCR artifacts using a mock community and human stool samples. It was
hypothesized that additional rounds of PCR would exacerbate the number
of artifacts. These PCR artifacts included (i) the effect of the
polymerase on the error rate of the bases represented in the final
sequences, (ii) the fraction of sequences that appeared to be chimeras
and the ability to detect those chimeras, (iii) the bias of
preferentially amplifying one fragment over another in a mixed pool of
templates, and (iv) inter-sample variation in community structure of
samples amplified with the same polymerase across the amplification
process. To characterize these factors we sequenced a mock community of
8 organisms with known sequences and community structure and human fecal
samples with unknown sequences and community structures. We sequenced
the V4 region of the 16S rRNA genes from a mock community by generating
paired 250 nt reads on the Illumina MiSeq platform. This region and
sequencing approach was used because it has been shown to result in a
relatively low sequencing error rate and is a widely used protocol (18).
To better understand the impact of DNA polymerase choice on PCR
artifacts, we selected five high fidelity DNA polymerases and amplified
the communities using 20, 25, 30, and 35 rounds of amplification.
Collectively, our results suggest that the number of rounds and to a
lesser extent the choice of DNA polymerase used in PCR impact the
sequence data. The effects are consistent and are smaller than the
biological differences between individuals.

@@

\textbf{\emph{The fraction of sequences identified as being chimeric
varies by the number of cycles and the DNA polymerase used in PCR.}}
Chimeric PCR products can significantly confound downstream analyses.
Although numerous bioinformatic tools exist to identify and remove
chimeric sequences with high specificity, their sensitivity is
relatively low and can be reduced by the presence of sequencing errors
(14, 35). Because the true sequences of the organisms in the mock
community were known, we generated all possible chimeras between pairs
of V4 16S rRNA gene fragments and used these possible chimeric sequences
to screen the sequences generated under the different PCR conditions for
chimeras. The number of chimeras increased with rounds of amplification
(Figure 2A). Interestingly, the fraction of chimeric sequences from the
mock community varied by the type of polymerase used. After 30 rounds of
PCR, the Platinum polymerase had the highest chimera rate (i.e.~18.2\%)
followed by the Q5 (i.e.~8.1\%), Phusion (i.e.~7.5\%), Kappa
(i.e.~2.3\%), and Accuprime (i.e.~0.9\%) polymerases. Because our
chimera screening procedure can only be applied to mock communities, we
used the UCHIME algorithm to model the chimera screening approach that
is used in most sequence curation pipelines. By comparing the output of
UCHIME to our reference-based approach, we were able to calculate the
UCHIME's sensitivity and specificity (Figure 2A). The specificity for
all polymerases was above 95.7\% and showed a weak association with the
number of cycles used (Figure 2A). There was considerable
inter-polymerase and inter-round of amplification variation in the
sensitivity of UCHIME to detect the chimeras from the mock community.
This suggested that the residual error rate after pre-clustering the
sequence data did not compromise the sensitivity of UCHIME to detect
chimeras. The sensitivity of UCHIME varied between 50 and 87.0\% when at
least 25 cycles were used. The generalizability of these results is
limited because we used a single mock community with limited genetic
diversity. Although we did not know the true chimera rate for our four
human stool samples, we were able to calculate the fraction of sequences
that UCHIME identified as being chimeric (Figure 2B). These results
followed those from the mock communities: additional rounds of
amplification significantly increased the rate of chimeras and there was
variation between the polymerases that we used. Although it was not
possible to identify the features of a polymerase that resulted in
higher rates of chimeras, it is clear that using the smallest number of
PCR cycles possible will minimize the impact of chimeras.

\textbf{\emph{At the sequence level, PCR amplification bias is subtle.}}
Since researchers began using PCR to amplify 16S rRNA gene fragments
there has been concern that amplifying fragments from a mixed template
pool could lead to a biased representation in the pool of products and
would confound downstream analyses (24, 27--33). The mock community was
generated by mixing equal amounts of genomic DNA from 8 bacteria
resulting in uneven representation of the \emph{rrn} operons across the
bacteria as each bacterium had a different genome size and varied in the
number of operons in its genome. The vendor of the mock community
subjects each lot of genomic DNA to shotgun sequencing to more
accurately quantify the actual abundance of each organism in the
community. We compared the expected relative abundance of the 16S rRNA
genes from each bacterium in the mock community to the data we generated
across rounds of amplification and polymerase (Figure 3). Interestingly,
for some bacteria, their representation became less biased with
additional rounds of PCR (e.g.~\emph{L. fermentum}), while others became
more biased (e.g.~\emph{E. faecalis}), and others had little change
(e.g.~\emph{B. subtilis}). Contrary to prior reports (28), the
percentage of bases in the V4 region that were guanines or cytosines was
not predictive of the amount of bias. Across the strains there was no
variation in the length of their V4 regions and they each had the same
sequence in the region that the primers annealed. One of the bacteria
represented in the mock community, \emph{S. enterica}, had 6 identical
copies of the V4 region and 1 copy that differed from those by one
nucleotide. The dominant copy had a thymidine and the rare copy had a
guanine. We used the sequence data to calculate the ratio of the
dominant to rare variants from \emph{S. enterica} expecting a ratio near
6 (Figure S1). The Accuprime, Phusion, Platinum, and Q5 polymerases
converged to a ratio of 5.4; however, the ratio for the Kappa polymerase
varied between 6.1 and 6.8 for 25 to 35 rounds of PCR. Given the subtle
nature of the variation in the relative abundances of each 16S rRNA gene
fragment, it was not possible to create generalizable rules that would
explain the bias.

\textbf{\emph{At the community level, the effects of PCR amplification
bias grow with additional rounds of PCR.}} Because the variation in bias
between polymerases and across rounds of PCR could be artificially
inflated due to sequencing errors and chimeras, we analyzed the alpha
and beta diversity of the mock community data at different phases of the
sequence curation pipeline (Figure 4). First, we removed the chimeras
from the mock community data as described above and mapped the
individual reads to the OTUs that the 16S rRNA gene fragments would
cluster into if there were no sequencing errors. This gave us a
community distribution that reflected the distribution following PCR
without any artifacts. Although the richness did not change, the Shannon
diversity increased with the number of rounds of PCR for all polymerases
except the Kappa polymerase, for which the diversity decreased (Figure
4A). These data suggest that PCR had the effect of making the community
distribution more even than it was originally, except for the data
generated using the Kappa polymerase where the evenness decreased. Next,
we used the observed sequence errors, but removed chimeras using our
reference-based approach, and clustered the reads to OTUs. The richness
and diversity metrics trended higher with higher error rates and number
of rounds of PCR (Figure 4A). Finally, we used the observed sequence
data and the UCHIME algorithm to identify chimeras. Again, the richness
and diversity metrics trended higher with higher error rates and number
of rounds of PCR (Figure 4A). These comparisons demonstrated that
although the bias at the sequence level was subtle, PCR introduces bias
at the community level that is exacerbated by errors and chimeras when
sequences are clustered into OTUs. When we measured the Bray-Curtis
distance between the communities observed after 25 rounds of
amplification and those at 30 and 35, distances between 25 and 35 rounds
were higher than between 25 and 30 rounds for each of the polymerases by
an average of 0.022 units (Figure 4B). The Platinum polymerase varied
the most across rounds of amplification (25 vs 30 rounds: 0.12; 25 vs 35
rounds: 0.15). Although the distances between samples were small, the
ordination of these distances showed a clear change in community
structure with increasing rounds of PCR (Figure 4C). This observation
was supported by our statistical analysis, which revealed that the
effect of the number of rounds of PCR (R\textsuperscript{2}=0.21,
P\textless{}0.001) was comparable to the choice of polymerase
(R\textsuperscript{2}=0.20, P\textless{}0.001). These results
demonstrate that subtle differences in relative abundances can have an
impact on overall community structure. This variation underscores the
importance of only comparing sequence data that have been generated
using the same PCR conditions.

\textbf{\emph{The choice of polymerase or the number of rounds of
amplification have little impact on the relative interpretation of
community-wide metrics of diversity.}} We expected that the biases that
we observed at the population and community levels using mock community
data would be small relative to the expected differences between
biological samples. To study this further, we calculated alpha and
beta-diversity metrics using the human stool samples for each of the
polymerases and rounds of amplification. We calculated the number of
observed OTUs and Shannon diversity for each condition and stool sample
(Figure 5A). Although there were clear differences between PCR
conditions, the relative ordering of the stool samples did not
meaningfully vary across conditions. When we characterized the variation
between rounds of amplification using human stool samples, the distance
between the 25 and 30 rounds and 25 and 35 rounds varied considerably
between samples and polymerases (Figure 5B). In general the inter-round
variation was lowest for the data generated using the Kappa and
Accuprime polymerases. The data generated using the Platinum polymerase
was consistent across rounds, but overall, it was more biased than the
other polymerases. Considering the average distance across the four
samples varied between 0.39 and 0.56, regardless of the polymerases and
number of rounds of amplification, any bias due to amplification is
unlikely to obscure community-wide differences between samples. In
support of this was our principle coordinates analysis of the
Bray-Curtis distances, which revealed distinct clusters by stool sample
(Figure 5C). Within each cluster there were no obvious patterns related
to the polymerase or number of rounds of PCR. However, our statistical
analysis revealed statistically significant differences in the community
structures with the stool sample explaining the most variation
(R\textsuperscript{2}=0.79, P\textless{}0.001), followed by the number
of rounds of PCR (R\textsuperscript{2}=0.044, P\textless{}0.001) and the
choice of polymerase (R\textsuperscript{2}=0.033, P\textless{}0.001).
Together, these results indicate that for a coarse analysis of
communities, the choice of number of rounds of amplification or
polymerase are not important, but that they must be consistent across
samples. It is difficult to develop a specific recommendation based on
the level of bias across rounds of PCR or polymerases; however, the
general suggestion is to use as few rounds of amplification as possible.

@@

Our results suggest that the number of rounds of PCR and to a lesser
degree the choice of DNA polymerase impact the analysis of 16S rRNA gene
sequence data from bacterial communities. Although it was not possible
to make direct connections between PCR conditions and specific sources
of bias, we were able to identify general recommendations that reduce
the amount of error, chimera formation, and bias. Researchers should
strive to minimize the number of rounds of PCR and should use a high
fidelity polymerase. Although specific PCR conditions impact the precise
interpretation of the data, the effects were consistent and were smaller
than the biological differences between the samples we tested. Based on
these observations, amplicons must be generated by consistent protocols
to yield meaningful comparisons. When comparing across studies, values
like richness, diversity, and relative abundances must be made in
relative and not absolute terms.

The observed sequencing error rates and alpha diversity metrics followed
the manufacturers' measurements of their polymerases' fidelity (Figure
1). Accuprime and Platinum have fidelity that are approximately 10-times
higher than that of Taq whereas the fidelity of Phusion, Q5, and Kappa
are more than 100 times higher. Among these polymerases, the Kappa
polymerase resulted in the the lowest error rate, lowest chimera rate,
and least bias across rounds of PCR. Considering polymerase development
is an active area of commercial development with potential new
polymerases becoming available, it is important for researchers to
understand how changing the polymerase impacts downstream analyses for
their type of samples.

Over the past 20 years, a large literature has attempted to document
various PCR biases and underscored the fact that data based on
amplification of DNA from a mixed community are not a true
representation of the actual community. In addition to obvious biases
imposed by primer selection, other factors inherent in PCR can influence
the representation of communities. Factors that can lead to preferential
amplification of one fragment over another have included guanine and
cytosine composition, length, flanking DNA composition, amount of DNA
shearing, and number of rounds of PCR (24, 27--33). In addition,
environmental and reagent contaminants can also have a significant
impact on the analysis of low biomass samples (39). Less well understood
is the effect of phylogenetic diversity on bias and chimera formation.
Communities with low phylogenetic diversity may be more prone to chimera
formation since chimeras are more likely to form among closely related
sequences (14, 35). The interaction of these various influences on PCR
artifacts are complex and difficult to tease apart. Minimizing the level
of DNA shearing and using the fewest number of rounds of PCR with a
polymerase that has the highest possible fidelity are strategies that
can be employed to minimize the formation of chimeras. Although care
should always be taken when choosing a polymerase for 16S rRNA gene
sequencing, our observations show that variation among polymerases is
smaller than the actual biological variation in fecal communities
between individuals.

Even with these strategies it is impossible to remove all PCR artifacts.
Beyond the imperfections of the best polymerases, sometimes difficult to
lyse organisms require stringent lysis steps and low biomass samples
require additional rounds of PCR. A host of bioinformatics tools are
available for removing residual sequencing errors (18, 40--42). These
tools struggle to correctly differentiate true biological diversity
(i.e.~Amplicon Sequencing Variants or ASVs) and PCR or sequencing
errors. Although these methods remove residual errors, they also risk
splitting intragenomic variants into separate ASVs or merging 16S rRNA
gene sequences from different taxa into the same ASV. Other tools are
available for removing chimeras (14, 35) where there is a trade off
between the sensitivity of detecting chimeras and the specificity of
correctly calling a sequence a chimera. In recent years, parameters for
these algorithms have been changed to increase their sensitivity with
little evaluation of the effects on the specificity of the algorithms
(40, 42). Others recommend removing any read that has an abundance below
a specified threshold as a tool to remove PCR and sequencing artifacts
(e.g.~removing all sequences that only appear once) (20, 40--42). This
method must be approached with caution as such approaches are likely to
introduce a different bias of the community representation and ignore
the fact that artifacts may be quite abundant. Ultimately, researchers
must test their hypotheses with multiple methods to validate the claims
they reach with any one method (43). All methods have biases and
limitations and we must use complementary methods to obtain robust
results.

@@

\textbf{\emph{PCR protocol.}} Five high fidelity DNA polymerases were
tested including AccuPrime\textsuperscript{TM} (ThermoFisher, MA, USA),
KAPA HIFI (Roche, IN, USA), Phusion (New England Biolabs, MA, USA),
Platinum (ThermoFisher, MA, USA), and Q5 (New England Biolabs, MA, USA).
Manufacturer recommendations were followed except for the annealing and
extension times, which were selected based on previously published
protocols (18, 37). Primers targeting the V4 region of the 16S rRNA gene
were used with modifications to generate MiSeq amplicon libraries (18)
(\url{https://github.com/SchlossLab/MiSeq_WetLab_SOP/}). The number of
rounds of PCR used for each sample and polymerase started at 15 and
increased by 5 rounds up to 35 cycles. Insufficient PCR product was
generated using 15 rounds and has not been included in our analysis.

\textbf{\emph{Library generation and sequencing.}} Each PCR condition
(i.e.~combination of polymerase and number of rounds of PCR) were
replicated four times for the mock community and one time for each fecal
sample. Libraries were generated as previously described (18)
(\url{https://github.com/SchlossLab/MiSeq_WetLab_SOP/}). The libraries
were sequenced using the Illumina MiSeq sequencing platform to generate
paired 250-nt reads.
